---
title: "Visualising the Pandemic"
subtitle: "Comparing trajectories of COVID-19 across countries "
author: "Andriy Koval"
date: "June 2, 2020"
output:
  html_document:
    theme: simplex
    toc: true
    toc_depth: 3
    keep_md: true
    toc_float: true
    code_folding: show
---

<!-- These two chunks should be added in the beginning of every .Rmd that you want to source an .R script -->
<!--  The 1st mandatory chunck  -->
<!--  Set the working directory to the repository's base directory -->
```{r, echo=F, message=F} 
#Don't combine this call with any other chunk 
# cat("Working directory: ", getwd()) # check where you are
  library(knitr)
# Rmd is in "./reports/reportA/reportA.Rmd", but now the root is "./"
  knitr::opts_knit$set(root.dir='../../') # "go up" from the folder where this .Rmd is save until you arrive at the root/working directory
```

<!--  The 2nd mandatory chunck  -->
<!-- Set the report-wide options, and point to the external code file. -->
```{r set_options, echo=F}
# set options shared by all chunks
opts_chunk$set(
  results     = 'show', 
  attr.source = ".numberLines",
  message     = TRUE,
  comment     = NA, 
  tidy        = FALSE,
  fig.height  = 4, 
  fig.width   = 5.5, 
  out.width   = "900px",
  fig.path    = 'figure_rmd/'     
)
echoChunks <- TRUE
options(width=120) #So the output is 50% wider than the default.
# connect to the file with the underlying R script  
read_chunk("./analysis/covid-trajectory/covid-trajectory-1.R")
# read_chunk("../../../analysis/blogposts/florida-suicide-count/fl-suicide-count.R")
```
# TL;DR

This workshop demonstrates basic exploration of COVID-19 data, with the focus on computing and graphing data on ___relative___ timelines, in which the "day zero" is unique for each country. 

# Learning Objectives

After this workshop participants should be able to:
 
1. Plot time series of COVID-19 cases using `ggplot2` package
2. Add interactive highlights to trajectories using `plotly` package
3. Compute indicators for key epidemiological events in each country (e.g. day of the first death)
4. Construct country-specific timelines relative to key epidemiological events
5. Visualize the sequence of key events for a group of countries

# Visualization Goals

In this tutorial we will have three visualisation goals. First, we would like to have a tool for plotting multiple trajectories onto the same canvas and explore the differences between trajectories plotted on the same scale.  Second, we would like to create a unique temporal context for each trajectory and express it with respect meaningful anchors, such as first detected case or first confirmed death. Thirdly, we would like to better understand the sequence of these temporal anchors to have a better understanding of how the pandemic unfolded. 

Here are the graphs we will be creating in this tutorial:  


```{r preview-0, echo=FALSE,  message=FALSE, results='show',warning = F, cache=TRUE}
```

## Goal 1 - Absolute, Interactive   
Create a line graph of many trajectories with an interactive tooltip.  
```{r preview-1, echo=FALSE,  message=FALSE, results='show',warning = F, cache=TRUE, fig.height = 6}
```

## Goal 2 - Relative, faceted 
Create a trajectory graph depicting how the epidemic unfolded in each contry relative to its key epidemiological events.  
```{r preview-2, echo=F, fig.height=4.5, fig.width=6.5, out.width = "900px", message=FALSE, results='asis',warning = F, cache=TRUE}
```


## Goal 3 - Temporal sequence
Create a graph depicting the sequence of key epidemiological events across countries.   
```{r preview-3, echo=F, fig.height=6, fig.width=8, out.width = "900px", message=FALSE, results='show',warning = F, cache=TRUE}
```


# Environment

Non-technical readers are welcome to skip this section. 
```{r load-packages, echo=echoChunks, message=FALSE}
```

```{r load-sources, echo=echoChunks, message=FALSE}
```

```{r declare-globals, echo=echoChunks, results='show', message=FALSE}
```
 

# Data  

The data comes from [European Centre for Disease Prevention and Control](https://www.ecdc.europa.eu/en), with the source available from [here](https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide). I demonstrate the preparation of this data for analysis in  [./manipulation/ellis-covid.R](https://github.com/andkov/vada-2020-summer-school/blob/master/manipulation/ellis-covid.R) script of this repository. 

```{r load-data, echo=echoChunks, results='show', message=FALSE}
```

A few things to notice about this data set:

1. Only `date` , `n_cases`, and `n_deaths` change over time. All other variables have a single unique value for each country.

2. For some countries, the observations are missing for certain dates, for example:
```{r}
ds_covid %>% filter(country_code == "FIN") %>% filter(date > as_date("2020-02-26"))
```
The data preparation step ensured that each country has the same number of rows, creating these missing cells. This is important for two reasons: __A__) we want to be able to differentiate between the absence of ___cases___ and the absence of ___reporting___ and __B__) missing dates will complicate comutation of relative timelines. 


# Data Tweaks

```{r tweak-data-1, echo=echoChunks, results='show', message=FALSE,cache=TRUE, warning =F}
```

We focus on 36 country members of the Organization for Economic Co-operation and Development (OECD) because they report  more nuanced data on their economic and social development, so we can have a richer pool of explanatory variables (see http://stats.oecd.org/ )


# Goal 1

 [plotly](https://plotly.com/) is a JavaScript library used for creating interactive visualizations and dashboards. Its implementation in `python` and `R`, the latter offered via `plotly` package. Its syntax is very similar to `ggplot2`, but nevertheless sufficiently different to be able to distract you into a (potentially lenghty) side quest. Luckily, there is a magic function `plotly::ggplotly` that transforms a graph built in `ggplot2` into an interactive `plotly` object. 
 
To demonstrate, let us first build a basic graph with multiple timeseries: 
```{r goal_1-1, echo=TRUE, message=FALSE, results='show',warning = F, cache=TRUE}
```

And now we enhanse this plot with  highlight component using the `plotly::highlight()` function and some additional options:
```{r goal_1-2, echo=TRUE, message=FALSE, results='show',warning = F, cache=TRUE, fig.height=6}
```


For more options and syntax guide, see https://plotly-r.com/client-side-linking.html. There is also a package for rendering `ggplot2` into interactive graphs ([`ggplotly`](https://rdrr.io/cran/plotly/man/ggplotly.html)), which offers less flexibility in display design compared to creating graphs directly with `plotly`, but is much simpler to implement. 


# Goal 2 

It is often makes sense to compare the progression of epidemics across countries using a meaningful "time zero", for example the day of the first confirmed case or the first confirmed death in the country. To create the Goal 2 graph we first need to derive a number of new variables. The following subgoals will help learn the operations used to derive them 

___Subgoals___:  

#### 1. Compute comulative cases (Running total)  

#### 2. Create onset markers (which values exceed a chosen threshold?)  

#### 3. First case marker (what row is the first to exceed a chosen threshold?)  

#### 4. Date of first case (what is the date associated with exceeding a chosen threshold?)  

#### 5. Number of days since 1st case (how many days have passed between current date and the date of exceeding a threshold?)  



## Reprex 

To help us carry out the computation, let us construct a fictional example that we can use to develop the script
```{r reprex-1,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE}
```

#### 1. Compute comulative cases (Running total)
 
Computing cumulative (running) sum is easily accomplished with `cumsum` function paired with `group_by`, however, watch  out for `NA` values: they will break the running sum, resulting in the `NA` for the rest of the column. To avoid this, we use convert `NA` to `0` on the fly with `tidyr::replace_na()` function

```{r reprex-2,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```

This option is better than converting `NA`s to zero during the data preparation step, as this would mask the absence of reporting, overwriting it with a definitive value of `0` cases. 

#### 2. Create onset markers (which values exceed a chosen threshold?)  

Now we will use a simple logical test to create a logical variable indicating when the running total exceeded the value of the chosen threshold. 
```{r reprex-3,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```
It may make sense to use other operationalization the "onset" event.. For example, we can define it as "the date of the 10th case" or ".01% of population infected" or "5th days of registering new cases in a row".  

#### 3. First case marker (what row is the first to exceed a chosen threshold?)  

Now with the column `onset_case` marking whether the running total is higher than a threshold, we can identify the first occurrence of `TRUE` : 
```{r reprex-4,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```
Notice, that we use the property of `logical` class: when used in mathematical expression, `FALSE` assumes the value of `0`, while `TRUE` is interpreted as `1`. 

#### 4. Date of first case (what is the date associated with exceeding a chosen threshold?)  

Now we can use this indicator to extract the date associated with this row:
```{r reprex-5,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```

There will be only one date associated with traspassing the threshold, so we populate the rest of the cells with it
```{r reprex-6,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```

#### 5. Number of days since 1st case (how many days have passed between current date and the date of exceeding a threshold?)  

This allows us for a very straightforward computation of the distance between any given date and the date of the "onset", in this case the date of the first confirmed case:
```{r reprex-7,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```

#### Re-write for brevity  
Finally, we can re-express these steps more succinctly, however, it might be advisible to leave these step in comments in case you need to retrace your steps or debug an error down the stream

```{r reprex-8,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```

## Compute relative timelines

The `reprex` data set we have created allows to inspect the logic of our script manually and have a much higher confidence that it does what we expect it to do. Now we can apply this solution to the real data. Notice that we create a variation of this solution, applying another definition of the "day zero" - the day of the first confirmed mortality in the country. 

```{r compute-epi-timeline,echo = echoChunks, message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```
Also notice that we create a few other helper variables: two other operationalization of the absolute time line (`days_since_exodus` and `days_since_pandemic`) as well as population adjustments for cumulative cases and deaths in each country(`n_cases_cum_per_1m` and `n_deaths_cum_per_1m`)


## Goal 2 graph 

With all variable prepared for plotting, let us develop the barebones graph first, to help us see the composition of the plot: 
```{r goal_2-1, echo=T, fig.height=4, fig.width=6, out.width = "900px", message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```

Now let's enhance the appearance of the plot:  
```{r goal_2-2, echo=T, fig.height=4, fig.width=6, out.width = "900px", message=FALSE, results='show',warning = F, cache=TRUE, attr.source='.numberLines'}
```

# Goal 3
Similarly with the plot from Goal 2, let us first sketch the essential components
```{r goal_3-1, echo=T, fig.height=6, fig.width=8, out.width = "900px", message=FALSE, results='show',warning = F, cache=TRUE}
```
and then enhance the appearance:  
```{r goal_3-2, echo=T, fig.height=6, fig.width=8, out.width = "900px", message=FALSE, results='show',warning = F, cache=TRUE}
```

# Exercises

To help you further you mastery of these graphical forms consider the following exercises:  

1. Using the graph from Goal 1, re-design the interactive highlights so that they depicts both the trajectory line and the dots for temporal anchors ( day of first case and day of first death). Hint: you can no longer filter within the geom, so you need to create a new variable __before__ tranforming the dataset with `highlight_key`.    

2. Re-draw the graph in Goal 2 with a different operationalization of the relative timeline: Days since 5 daily new deaths first reported.  

3. Redraw the graph in Goal 3 to sort the countries according to the descending distance between the day of the first case and the day of the first death.   


session information
===========================================================================

For the sake of documentation and reproducibility, the current report was rendered in the following environment.  Click the line below to expand.

<details>
  <summary>Environment <span class="glyphicon glyphicon-plus-sign"></span></summary>
```{r session-info, echo=FALSE}
if( requireNamespace("devtools", quietly = TRUE) ) {
  devtools::session_info()
} else {
  sessionInfo()
} 
```
